---
title: "Final Project - Practical Machine Learning"
author: "Umair Rafique"
date: "May 19, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

The data for this project come from this source: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har

This was the training dataset:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

And this was the testing dataset:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The goal of the project will be to predict the manner in which the subjects did the exercise, which is the 'classe' variable in the training data.

## Getting and Cleaning the data

```{r message=FALSE}
df_trainData <- read.csv("pml-training.csv")
df_testData <- read.csv("pml-testing.csv")
library(caret)
library(dplyr)
```

We will first explore how much valid information each column contains, to decide which ones to retain for training our model

```{r}
colMeans(is.na(df_trainData))
```

We will get rid of the columns that have > 90% NA's

```{r}
varsToDrop <- names(colMeans(is.na(df_trainData)))[colMeans(is.na(df_trainData)) > 0.9]

df_trainData %>% 
  select(-varsToDrop) -> df_trainData
df_testData %>% 
  select(-varsToDrop) -> df_testData
```

We will also drop the columns that have near-zero variance, as they don't contain information that will help us in predicting the outcome. We will also drop the first column which contains the row numbers.

```{r}
library(caret)

vars_nzv <- nearZeroVar(df_trainData, saveMetrics = TRUE)$nzv

df_trainData <- df_trainData[, vars_nzv==FALSE]
df_testData <- df_testData[, vars_nzv==FALSE]

df_trainData <- df_trainData[,-1]
df_testData <- df_testData[,-1]
```

## Training our model

To train our model, we will first split the training data into two sets: Training, and cross-validation, and will name them train and test, respectively. After that we will first train a random forest on our data, followed by a xgboost, and see which one performs best.

```{r}
inTrain <- createDataPartition(df_trainData$classe, p = 0.7, list = FALSE)
train <- df_trainData[inTrain, ]
test <- df_trainData[-inTrain, ]
```

### Random Forest
```{r}
mdl_rf <- train(classe ~ ., method = "rf", data = train)
mdl_rf
confusionMatrix(test$classe, predict(mdl_rf, test))
```

### XGBoost
```{r}
mdl_xgb <- train(classe ~ ., method = "xgboost", data = train)
mdl_xgb
confusionMatrix(test$classe, predict(mdl_xgb, test))
```


